\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2021-11-28 main.Rmd
editing before creating a cubble in Section 2

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path = "figures/",
  fig.height = 10
)
options(prompt = "R> ", continue = "+ ")
```

@@

Many data structures have been proposed for spatial (`sf` by @sf) and temporal (`tsibble` by @tsibble) data in the R community, while less has been done for spatio-temporal data. The lack of such tools could potentially  because analysts usually treat the spatial and temporal dimension pf the data separately, without realising the need to create a new data structure. While this approach follows the third tidy data principal [@tidydata] (*Each type of observational unit forms a table*), analysts always need to manually join results from different observational units or combining multiple tables into one for downstream analysis. This additional step doesn't add new operations into the data but can be error prone.  \newline

@@

Spatio-temporal data usually come in various forms and Figure \ref{fig:cubble-diagram} shows four examples of this. No matter whichever form the data is in, there are always some common components shared by these data. A spatial identifier (`id` in the diagram) identifies each site. The temporal identifier (`t` in the diagram) [...]. Coordinates, comprising of latitude and longitude, are commonly used variables for point pattern data.These identifiers will be the building blocks for the data structure introduced below. For other variables, those invariant at each time stamp are spatial variables and those differ are temporal variables.

@@

In a cubble, there are two forms: 1) nested form, for manipulating the spatial dimension, and 2) long form, for manipulating the temporal dimension. Figure \ref{fig:def-cubble} sketches the two forms along with the associated attributes. A variable identifies by the spatial identifies can come from manipulating spatial variables itself, or summary of temporal variables. The nested cubble is best suited to work with this type of operation, since it defines each spatial unit as a row. The spatial variables are directly displayed in columns. Temporal variables are nested in a column called `ts` and the underlying rowwise dataframe uses a `group` attributes to ensure each row  is in its own group.

<!-- In this structure, the nested form is suitable for manipulating the spatial dimension of the data and this would include manipulating variables that 1) are inherently spatial and 2) summarise the temporal variables by `key`. Under the hood, the nested cubble is built on top of the rowwise dataframe (`rowwise_df`). This is design to simplify the code when working with the temporal variables - they are nested into the list-column `ts`. -->

Temporal operations are suited to be performed in the long cubble as each row is defines as the combination of spatial and temporal identifier. This is also the structure that `tsibble` adopts. Temporal variables are directly displayed. To avoid repeating the same spatial at each temporal unit, all the spatial variables, along with the spatial identifier, are stored as a `spatial` attributes. This information is used when switching back to the nested form.

@@

To use functionalities from cubble, data analysts first need to create a cubble. `as_cubble` create a `cubble` by supplying the three key components: `key` as the spatial identifier; `index` as the temporal identifier; and a vector of `coords` in the order of longitude first and then latitude. The use of `key` and `index` follows the naming convention in `tsibble`. The cubble created by default is in the nested form. Below is an example of creating a cubble:  \newline

@@

In the `cubble` header, you can read the name of the `key` variable, bbox, and also the name of variable nested in the `ts` column. In this example, the spatial identifier is `id` and the number in the bracket means there are 5 unique `id` in this dataset. The bbox in the second row gives the range of the coordinates. The temporal variables are all nested in the `ts` column, but it could be useful to know the name these variables. The third row in the cubble header shows these names and in this example this includes: precipitation, `prcp`, maximum temporature, `tmax`, and minimum temperature, `tmin`.

@@

`cubble` is also compatible with `tsibble` in the sense that the original list-column can be a `tbl_ts` object. Duplicates and gaps shoudl be first checked before structuring the data into a cubble. If the input data is a `tsibble` object, the long form cubble is also a `tsibble` where users can directly apply time series operations.

@@

In some visualisation, we may need information from both spatial and temporal dimension. One example of this is a glyph map, where spatial variables, i.e. `longitude` and `latitude`, are used to construct the major axes and temporal variables, i.e. `month` and `tmax`, are used to construct the minor axes. This requires these variables to be in the same table, rather than in different forms. In cubble, you can append spatial variables that are invariant to the `key` withe `migrate()`

@@

Imposing a clustering structure can be thought of as building a hierarchical structure where stations are nested within clusters. As an example to illustrate here, we use a kmean clustering algorithm based on the distance matrix and specify the number of centers to be 20. More complex algorithms can also be used for more complex problem, as long as a mapping from each station id to the cluster id can be constructed. We then join this data to our station data:

@@

With a cubic framework on mind, different types of manipulation with cubble can be thought of as slicing the cube in various way. The table below shows how some `dplyr` verbs are mapped into the operation in a cubble. With the existing grouping on the station, additional groupping can be added with `group_by` and removed with `ungrouped`. [talk about why it is useful]

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  fig.path = "figures/",
  fig.height = 10
)
options(prompt = "R> ", continue = "+ ")
```

@@

Many data structures have been proposed for spatial (`sf` by @sf) and temporal (`tsibble` by @tsibble) data in the R community, while less has been done for spatio-temporal data. The lack of such tools could potentially  because analysts usually treat the spatial and temporal dimension of the data separately, without realising the need to create a new data structure. While this approach follows the third tidy data principal [@tidydata] (*Each type of observational unit forms a table*), analysts always need to manually join results from different observational units or combining multiple tables into one for downstream analysis. This additional step doesn't add new operations into the data but can be error prone.  \newline

@@

Spatio-temporal data usually come in various forms and Figure \ref{fig:cubble-diagram} shows four examples of this. No matter which form the data is in, these formats share some common components that characterise spatio-temporal data. A spatial identifier (`id` in the diagram) is the unique identifier of each site. The temporal identifier (`t` in the diagram) prescribes the time stamp each site is recorded. Coordinates, comprising of latitude and longitude (`lon` and `lat` in the diagram), locates each site on the map. These identifiers will be the building blocks for the data structure introduced below. Other variables in the data can be categorised into two groups: spatial variables that are invariant at each time stamp for every site, i.e. the name or code of the weather station and temporal variables that varies with time.

@@

In a cubble, there are two forms: nested form and long form, and Figure \ref{fig:def-cubble} sketches the two forms along with the associated attributes. The decision on which form to use is output-oriented, meaning analysts need to first think about whether the output of a particular operation is identified only by the spatial identifier, or a combination of spatial and temporal identifier. The nested cubble is suitable for working with operations that are only identified by site and this type of operation can be a pure manipulation of spatial variables, or a summary of temporal variables by site (i.e. the output of counting the number of raining day is only identified by sites and hence should be performed with the nested form). Underneath the nested form, a cubble is built from a row-wise dataframe (`rowwise_df`) where each site forms a separate group. This structure simplifies the calculation that involves temporal variables by avoiding the use of `map` syntax when working with list-column.

For those operations whose output involves both a spatial and temporal dimension, long form should be used. The long form is identified by both the spatial and temporal identifier and adopts a grouped dataframe (`grouped_df`) to forms each site as a group. Spatial variables are stored separately in a `tibble` as an special attribute of the long cubble. This design avoids repeating the spatial variables at each time stamp while not dropping information from spatial variables.

@@

To use functionalities from cubble, data analysts first need to create a cubble. `as_cubble` create a `cubble` by supplying the three key components: `key` as the spatial identifier; `index` as the temporal identifier; and a vector of `coords` in the order of longitude first and then latitude. The naming of `key` and `index` follows the convention in the `tsibble` package. The cubble created by default is in the nested form. Below is an example of creating a cubble:  \newline

@@

In the `cubble` header, you can read the name of the `key` variable, bbox, and also the name of variable nested in the `ts` column. In this example, the spatial identifier is `id` and the number in the bracket means there are 5 unique `id` in this dataset. The bbox in the second row gives the range of the coordinates. The temporal variables are all nested in the `ts` column, but it could be useful to know the name these variables. The third row in the cubble header shows these names and in this example this includes: precipitation, `prcp`, maximum temperature, `tmax`, and minimum temperature, `tmin`.

@@

`cubble` is also compatible with `tsibble` in the sense that the original list-column can be a `tbl_ts` object. Duplicates and gaps should be first checked before structuring the data into a cubble. If the input data is a `tsibble` object, the long form cubble is also a `tsibble` where users can directly apply time series operations.

@@

In some visualisation, we may need information from both spatial and temporal dimension. One example of this is a glyph map, where spatial variables, i.e. `longitude` and `latitude`, are used to construct the major axes and temporal variables, i.e. `month` and `tmax`, are used to construct the minor axes. This requires these variables to be in the same table, rather than in different forms. In cubble, you can append spatial variables that are invariant to the `key` with `migrate()`

@@

Imposing a clustering structure can be thought of as building a hierarchical structure where stations are nested within clusters. As an example to illustrate here, we use a kmean clustering algorithm based on the distance matrix and specify the number of centres to be 20. More complex algorithms can also be used for more complex problem, as long as a mapping from each station id to the cluster id can be constructed. We then join this data to our station data:

@@

With a cubic framework on mind, different types of manipulation with cubble can be thought of as slicing the cube in various way. The table below shows how some `dplyr` verbs are mapped into the operation in a cubble. With the existing grouping on the station, additional grouping can be added with `group_by` and removed with `ungrouped`. [talk about why it is useful]

### 2021-11-28 main.tex
editing before creating a cubble in Section 2

Many data structures have been proposed for spatial (\texttt{sf} by
\citet{sf}) and temporal (\texttt{tsibble} by \citet{tsibble}) data in
the R community, while less has been done for spatio-temporal data. The
lack of such tools could potentially because analysts usually treat the
spatial and temporal dimension pf the data separately, without realising
the need to create a new data structure. While this approach follows the
third tidy data principal \citep{tidydata} (\emph{Each type of
observational unit forms a table}), analysts always need to manually
join results from different observational units or combining multiple
tables into one for downstream analysis. This additional step doesn't
add new operations into the data but can be error prone. \newline

@@

Spatio-temporal data usually come in various forms and Figure
\ref{fig:cubble-diagram} shows four examples of this. No matter
whichever form the data is in, there are always some common components
shared by these data. A spatial identifier (\texttt{id} in the diagram)
identifies each site. The temporal identifier (\texttt{t} in the
diagram) {[}\ldots{]}. Coordinates, comprising of latitude and
longitude, are commonly used variables for point pattern data.These
identifiers will be the building blocks for the data structure
introduced below. For other variables, those invariant at each time
stamp are spatial variables and those differ are temporal variables.

@@

In a cubble, there are two forms: 1) nested form, for manipulating the
spatial dimension, and 2) long form, for manipulating the temporal
dimension. Figure \ref{fig:def-cubble} sketches the two forms along with
the associated attributes. A variable identifies by the spatial
identifies can come from manipulating spatial variables itself, or
summary of temporal variables. The nested cubble is best suited to work
with this type of operation, since it defines each spatial unit as a
row. The spatial variables are directly displayed in columns. Temporal
variables are nested in a column called \texttt{ts} and the underlying
rowwise dataframe uses a \texttt{group} attributes to ensure each row is
in its own group.

Temporal operations are suited to be performed in the long cubble as
each row is defines as the combination of spatial and temporal
identifier. This is also the structure that \texttt{tsibble} adopts.
Temporal variables are directly displayed. To avoid repeating the same
spatial at each temporal unit, all the spatial variables, along with the
spatial identifier, are stored as a \texttt{spatial} attributes. This
information is used when switching back to the nested form.

@@

To use functionalities from cubble, data analysts first need to create a
cubble. \texttt{as\_cubble} create a \texttt{cubble} by supplying the
three key components: \texttt{key} as the spatial identifier;
\texttt{index} as the temporal identifier; and a vector of
\texttt{coords} in the order of longitude first and then latitude. The
use of \texttt{key} and \texttt{index} follows the naming convention in
\texttt{tsibble}. The cubble created by default is in the nested form.
Below is an example of creating a cubble: \newline

@@

In the \texttt{cubble} header, you can read the name of the \texttt{key}
variable, bbox, and also the name of variable nested in the \texttt{ts}
column. In this example, the spatial identifier is \texttt{id} and the
number in the bracket means there are 5 unique \texttt{id} in this
dataset. The bbox in the second row gives the range of the coordinates.
The temporal variables are all nested in the \texttt{ts} column, but it
could be useful to know the name these variables. The third row in the
cubble header shows these names and in this example this includes:
precipitation, \texttt{prcp}, maximum temporature, \texttt{tmax}, and
minimum temperature, \texttt{tmin}.

@@

\texttt{cubble} is also compatible with \texttt{tsibble} in the sense
that the original list-column can be a \texttt{tbl\_ts} object.
Duplicates and gaps shoudl be first checked before structuring the data
into a cubble. If the input data is a \texttt{tsibble} object, the long
form cubble is also a \texttt{tsibble} where users can directly apply
time series operations.

@@

In some visualisation, we may need information from both spatial and
temporal dimension. One example of this is a glyph map, where spatial
variables, i.e.~\texttt{longitude} and \texttt{latitude}, are used to
construct the major axes and temporal variables, i.e.~\texttt{month} and
\texttt{tmax}, are used to construct the minor axes. This requires these
variables to be in the same table, rather than in different forms. In
cubble, you can append spatial variables that are invariant to the
\texttt{key} withe \texttt{migrate()}

@@

With a cubic framework on mind, different types of manipulation with
cubble can be thought of as slicing the cube in various way. The table
below shows how some \texttt{dplyr} verbs are mapped into the operation
in a cubble. With the existing grouping on the station, additional
groupping can be added with \texttt{group\_by} and removed with
\texttt{ungrouped}. {[}talk about why it is useful{]}

>>>>

Many data structures have been proposed for spatial (\texttt{sf} by
\citet{sf}) and temporal (\texttt{tsibble} by \citet{tsibble}) data in
the R community, while less has been done for spatio-temporal data. The
lack of such tools could potentially because analysts usually treat the
spatial and temporal dimension of the data separately, without realising
the need to create a new data structure. While this approach follows the
third tidy data principal \citep{tidydata} (\emph{Each type of
observational unit forms a table}), analysts always need to manually
join results from different observational units or combining multiple
tables into one for downstream analysis. This additional step doesn't
add new operations into the data but can be error prone. \newline

@@

Spatio-temporal data usually come in various forms and Figure
\ref{fig:cubble-diagram} shows four examples of this. No matter which
form the data is in, these formats share some common components that
characterise spatio-temporal data. A spatial identifier (\texttt{id} in
the diagram) is the unique identifier of each site. The temporal
identifier (\texttt{t} in the diagram) prescribes the time stamp each
site is recorded. Coordinates, comprising of latitude and longitude
(\texttt{lon} and \texttt{lat} in the diagram), locates each site on the
map. These identifiers will be the building blocks for the data
structure introduced below. Other variables in the data can be
categorised into two groups: spatial variables that are invariant at
each time stamp for every site, i.e.~the name or code of the weather
station and temporal variables that varies with time.

@@

In a cubble, there are two forms: nested form and long form, and Figure
\ref{fig:def-cubble} sketches the two forms along with the associated
attributes. The decision on which form to use is output-oriented,
meaning analysts need to first think about whether the output of a
particular operation is identified only by the spatial identifier, or a
combination of spatial and temporal identifier. The nested cubble is
suitable for working with operations that are only identified by site
and this type of operation can be a pure manipulation of spatial
variables, or a summary of temporal variables by site (i.e.~the output
of counting the number of raining day is only identified by sites and
hence should be performed with the nested form). Underneath the nested
form, a cubble is built from a row-wise dataframe (\texttt{rowwise\_df})
where each site forms a separate group. This structure simplifies the
calculation that involves temporal variables by avoiding the use of
\texttt{map} syntax when working with list-column.

For those operations whose output involves both a spatial and temporal
dimension, long form should be used. The long form is identified by both
the spatial and temporal identifier and adopts a grouped dataframe
(\texttt{grouped\_df}) to forms each site as a group. Spatial variables
are stored separately in a \texttt{tibble} as an special attribute of
the long cubble. This design avoids repeating the spatial variables at
each time stamp while not dropping information from spatial variables.

@@

To use functionalities from cubble, data analysts first need to create a
cubble. \texttt{as\_cubble} create a \texttt{cubble} by supplying the
three key components: \texttt{key} as the spatial identifier;
\texttt{index} as the temporal identifier; and a vector of
\texttt{coords} in the order of longitude first and then latitude. The
naming of \texttt{key} and \texttt{index} follows the convention in the
\texttt{tsibble} package. The cubble created by default is in the nested
form. Below is an example of creating a cubble: \newline

@@

In the \texttt{cubble} header, you can read the name of the \texttt{key}
variable, bbox, and also the name of variable nested in the \texttt{ts}
column. In this example, the spatial identifier is \texttt{id} and the
number in the bracket means there are 5 unique \texttt{id} in this
dataset. The bbox in the second row gives the range of the coordinates.
The temporal variables are all nested in the \texttt{ts} column, but it
could be useful to know the name these variables. The third row in the
cubble header shows these names and in this example this includes:
precipitation, \texttt{prcp}, maximum temperature, \texttt{tmax}, and
minimum temperature, \texttt{tmin}.

@@

\texttt{cubble} is also compatible with \texttt{tsibble} in the sense
that the original list-column can be a \texttt{tbl\_ts} object.
Duplicates and gaps should be first checked before structuring the data
into a cubble. If the input data is a \texttt{tsibble} object, the long
form cubble is also a \texttt{tsibble} where users can directly apply
time series operations.

@@

In some visualisation, we may need information from both spatial and
temporal dimension. One example of this is a glyph map, where spatial
variables, i.e.~\texttt{longitude} and \texttt{latitude}, are used to
construct the major axes and temporal variables, i.e.~\texttt{month} and
\texttt{tmax}, are used to construct the minor axes. This requires these
variables to be in the same table, rather than in different forms. In
cubble, you can append spatial variables that are invariant to the
\texttt{key} with \texttt{migrate()}

@@

With a cubic framework on mind, different types of manipulation with
cubble can be thought of as slicing the cube in various way. The table
below shows how some \texttt{dplyr} verbs are mapped into the operation
in a cubble. With the existing grouping on the station, additional
grouping can be added with \texttt{group\_by} and removed with
\texttt{ungrouped}. {[}talk about why it is useful{]}
